\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-07-18 analysis/paper/paper.Rmd
cleaned cs1

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(tidyverse)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- TRUE
eval_cs3 <- TRUE
eval_cs4 <- TRUE
eval_colophon <- TRUE

@@

$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 X.$$

@@

$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X.$$

@@

$$ \log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. $$

@@

\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               

@@

The functions in the casebase package can be divided into two categories: 1) data visualization, in the form of population-time plots; and 2) parametric modeling. We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our unit tests.

@@

@@

For the first case study, we will use of the European Randomized Study of Prostate Cancer Screening data. This dataset is available through the \pkg{casebase} package:

```{r erspc-data, eval = eval_cs1}
data(ERSPC)
ERSPC$ScrArm <- factor(ERSPC$ScrArm,
  levels = c(0, 1),
  labels = c("Control group", "Screening group")
)
```

The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].

Population time plots can be extremely informative graphical displays of survival data. They should be the first step in an exploratory data analysis. We facilitate this task in the \pkg{casebase} package using the \code{popTime} function. We first create the necessary dataset for producing the population time plots, and we can generate the plot by using the corresponding \code{plot} method:

```{r plot-erspc-data, eval = eval_cs1}
pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
plot(pt_object)
```

We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC,
  event = "DeadOfPrCa",
  exposure = "ScrArm"
)
plot(pt_object_strat)
```

We can also plot them side-by-side using the \code{ncol} argument:

```{r plot-stratified-erspc-data-side, eval = eval_cs1}
plot(pt_object_strat, facet.params = list(ncol = 2))
```

ADD PARAGRAPH ABOUT WHAT WE CAN CONCLUDE FROM THESE GRAPHS.

Next, we investigate the differences between the control and the screening arms. A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group). In \proglang{R}, it can be done as follows:

```{r erspc-cox, eval = eval_cs1}
cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
  data = ERSPC
)
summary(cox_model)
```

We can also plot the cumulative incidence function (CIF) for each group. However, this involves estimating the baseline hazard, which Cox regression treated as a nuisance parameter.

```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code}
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group"),
  ignore = 99
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
legend("topleft",
  legend = c("Control group", "Screening group"),
  col = c("red", "blue"),
  lty = c(1, 1),
  bg = "gray90"
)
```

Next, we show how case-base sampling can be used to carry out a similar analysis. We fit several models that differ in how we choose to model time.

<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any). -->

First, we will fit an exponential model. Recall that this corresponds to excluding time from the linear predictor.

```{r erspc-casebase-exponential, eval = eval_cs1}
casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
  data = ERSPC,
  ratio = 100
)

summary(casebase_exponential)
exp(coef(casebase_exponential)[2])
exp(confint(casebase_exponential)[2, ])
```

<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->

<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->

We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile. In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:

```{r erspc-casebase-cif, eval = eval_cs1, echo = echo_plot_code}
smooth_risk_exp <- casebase::absoluteRisk(
  object = casebase_exponential,
  time = seq(0, 15, 0.1),
  newdata = new_data
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)

legend("topleft",
  legend = c(
    "Control group (Cox)", "Control group (Casebase)",
    "Screening group (Cox)", "Screening group (Casebase)"
  ),
  col = c("red", "red", "blue", "blue"),
  lty = c(1, 2, 1, 2),
  bg = "gray90"
)
```

As we can see, the exponential model gives us an estimate of the hazard ratio that is similar to the one obtained using Cox regression. However, the CIF estimates are quite different. Based on what we observed in the population time plot, where more events are observed later on in time, we do not expect a constant hazard would be a good description for this data. In other words, this poor fit for the exponential hazard is expected. A constant hazard model would overestimate the cumulative incidence earlier on in time, and underestimate it later on; this is what we see on the cumulative incidence plot. This example demonstrates the benefits of population time plots as an exploratory analysis tool.

For the next model, we include time as a linear term. Recall that this corresponds to a Weibull model.

```{r erspc-casebase-weibull, eval = eval_cs1}
casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm,
  data = ERSPC,
  ratio = 100
)

summary(casebase_time)
exp(coef(casebase_time))
exp(confint(casebase_time))
```

Again, the estimate of the hazard ratio is similar to that obtained from Cox regression. We then look at the estimate of the CIF:

```{r erspc-casebase-weibull-cif, eval = eval_cs1, echo = echo_plot_code}
smooth_risk_time <- casebase::absoluteRisk(
  object = casebase_time,
  time = seq(0, 15, 0.1),
  newdata = new_data
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
lines(smooth_risk_time[, 1], smooth_risk_time[, 2], col = "red", lty = 2)
lines(smooth_risk_time[, 1], smooth_risk_time[, 3], col = "blue", lty = 2)

legend("topleft",
  legend = c(
    "Control group (Cox)", "Control group (Casebase)",
    "Screening group (Cox)", "Screening group (Casebase)"
  ),
  col = c("red", "red", "blue", "blue"),
  lty = c(1, 2, 1, 2),
  bg = "gray90"
)
```

We see that the Weibull model leads to a better fit of the CIF than the Exponential model, when compared with the semi-parametric approach.

Finally, we can model the hazard as a smooth function of time using the \code{splines} package:

```{r erspc-casebase-splines, eval = eval_cs1}
casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm,
  data = ERSPC,
  ratio = 100
)

summary(casebase_splines)
exp(coef(casebase_splines))
exp(confint(casebase_splines))
```

Once again, we can see that the estimate of the hazard ratio is similar to that from the Cox regression. We then look at the estimate of the CIF:

```{r erspc-casebase-splines-cif, eval = eval_cs1, echo = echo_plot_code}
smooth_risk_splines <- absoluteRisk(
  object = casebase_splines,
  time = seq(0, 15, 0.1),
  newdata = new_data
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
lines(smooth_risk_splines[, 1], smooth_risk_splines[, 2], col = "red", lty = 2)
lines(smooth_risk_splines[, 1], smooth_risk_splines[, 3], col = "blue", lty = 2)

legend("topleft",
  legend = c(
    "Control group (Cox)", "Control group (Casebase)",
    "Screening group (Cox)", "Screening group (Casebase)"
  ),
  col = c("red", "red", "blue", "blue"),
  lty = c(1, 2, 1, 2),
  bg = "gray90"
)
```

Qualitatively, the combination of splines and case-base sampling produces the parametric model that most closely resembles the Cox model. In the following table, we see that the confidence intervals are also similar across all four models. In other words, this reinforces the idea that, under proportional hazards, we do not need to model the full hazard to obtain reliable estimates of the hazard ratio. Of course, different parametric models for the hazards give rise to qualitatively different estimates for the CIF.

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
CI_exp <- c(exp(coef(casebase_exponential))["ScrArmScreening group"], exp(confint(casebase_exponential))["ScrArmScreening group", ])
CI_time <- c(exp(coef(casebase_time))["ScrArmScreening group"], exp(confint(casebase_time))["ScrArmScreening group", ])
CI_splines <- c(exp(coef(casebase_splines))["ScrArmScreening group"], exp(confint(casebase_splines))["ScrArmScreening group", ])

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE, 
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) # for fitting gompertz AFT
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(visreg)
library(lubridate)
library(tidyverse)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

eval_introduction <- TRUE
eval_theory <- TRUE
eval_implementation <- TRUE
eval_cs1 <- TRUE
eval_cs2 <- FALSE
eval_cs3 <- FALSE
eval_cs4 <- FALSE
eval_colophon <- TRUE

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 X. \label{eq:exp}
\end{equation}

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 t + \beta_2 X. \label{eq:gomp}
\end{equation}

@@

\begin{equation}
\log(\lambda(t; X)) = \beta_0 + \beta_1 \log(t) + \beta_2 X. \label{eq:weibull}
\end{equation}

@@

\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               \\

@@

The functions in the casebase package can be divided into two categories: 1) exploratory data analysis, in the form of population-time plots; and 2) parametric modeling of the hazard function. We explicitly aimed at being compatible with both \code{data.frame}s and \code{data.table}s. This is evident in some of the coding choices we made, and it is also reflected in our unit tests.

@@

The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:

1. The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional). This will create an object of class \code{popTime}.  
2. The corresponding \code{plot} method for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.

The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} [@ggplot2] family of functions for further customization.

To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening (ERSPC) data [@schroder2009screening] which was extracted using the approach described in [@liu2014recovering] and available through the \pkg{casebase} package. This data contains the individual observations for 159,893 men between the ages of 55 and 69 years at entry who were recruited from seven European countries.

We first create the necessary dataset for producing the population time plot using the \code{casebase::popTime} function. In this example we stratify by the treatment group. The resulting object inherits from class `popTime` and stores the exposure variable as an attribute:

```{r erspc-data, eval = eval_cs1, echo = c(-1:-3)}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))

pt_object <- casebase::popTime(ERSPC, 
                               time = "Follow.Up.Time",
                               event = "DeadOfPrCa",
                               exposure = "ScrArm")
inherits(pt_object, "popTime")
attr(pt_object, "exposure")
```

We then pass this object to the corresponding \code{plot} method which leverages the \pkg{ggplot2} package to build population time plots. 
```{r plot-stratified-erspc-data-code, eval = FALSE, echo = TRUE}
plot(pt_object, add.base.series = TRUE)
```

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.5,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray30"),
     case.params = list(size = 0.5),
     base.params = list(size = 0.5)) + theme_bw(base_size = 10) + theme(legend.position = "bottom") + 
  scale_y_continuous(labels = scales::label_number_si())
```

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by first adding a layer for the area representing the population time in gray, with subjects having the least amount of observation time plotted at the top of the y-axis. We immediately notice a unique _step shape_ in the population time area which is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999. This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the January 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is a combination of deaths and staggered entries.

Next, layers for the case series and base series are added. The case series is sampled at random vertically on the plot to avoid having all points along the upper edge of the plot, while the base series is sampled horizontally with sampling weight proportional to their follow-up time. By randomly distributing the points, we can get a better sense of the incidence density. In the ERSPC data, we see that more events are observed later on in time. Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on. The reader is referred to the package vignettes for more examples and a detailed description of how to modify the aesthetics of a population time plot.

@@

We investigate the differences in survival between the control and screening arms in the ERSPC study presented earlier using casebase sampling to fit a parametric hazard function. We fit four models that differ in the functional form of time; (1) excluded from the linear predictor (Eq.\eqref{eq:exp}), (2) linear function (Eq.\eqref{eq:gomp}), (3) log function (Eq.\eqref{eq:weibull}), and (4) a smooth function using cubic B-splines. The models are fit using \code{fitSmoothHazard}. Since the output object from \code{fitSmoothHazard} inherits from the \code{glm} class, we see a familiar result when using the function \code{summary}:

```{r, eval = FALSE, echo = TRUE}
fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(i) {
  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
})

summary(fits[[2]])
```

The \code{absoluteRisk} function takes as input the \code{fitSmoothHazard} object and returns a matrix where each column corresponds to the covariate profiles specified in the \code{newdata} argument, and each row corresponds to a specified time point:

```{r, eval=FALSE, echo=echo_plot_code}
new_data <- data.frame(ScrArm = c("Control group", "Screening group"))

risks <- lapply(fits, function(i) {
  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
})

head(risks[[2]])
```

In Figure \ref{fig:erspc-cox-cif}, we overlay the estimated CIFs from casebase on the Cox model CIF. The CIF estimates for the exponential model in panel (1) overestimate the cumulative incidence earlier on in time, and underestimate it later on. Based on what we observed in the population time plot, where more events are observed later on in time, this poor fit for the exponential hazard is expected. This example demonstrates the benefits of population time plots as an exploratory analysis tool.

In Figure \ref{fig:erspc-cox-cif} panels (2)-(4), we notice a better fit with increasing complexities of time. As noted above, the usual asymptotic results hold for likelihood ratio tests built using casebase sampling models. Therefore, we can easily test the null hypothesis that the exponential model is just as good as the larger (in terms of number of parameters) splines model:

```{r, eval = eval_cs1, echo=FALSE}
fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(i) {
  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
})
```

```{r erscp-compare, eval = eval_cs1, echo = FALSE}
anova(fits$exponential, fits$splines, test = "LRT")
```

As expected, we see that splines model provides a better fit. Similarly, we can compare the AIC for all four casebase models:

```{r erscp-aic, eval = eval_cs1, echo=FALSE}
c("Exp." = AIC(fits$exponential),
  "Gomp." = AIC(fits$gompertz),
  "Weib." = AIC(fits$exponential),
  "Splines" = AIC(fits$splines))
```

<!--A common choice for this type of analysis is to use a Cox regression model and estimate the hazard ratio for the screening group (relative to the control group). In \proglang{R}, it can be done as follows:-->

```{r erspc-cox, eval = eval_cs1, echo=FALSE}
surv_obj <- with(ERSPC, Surv(Follow.Up.Time, DeadOfPrCa))
cox_model <- survival::coxph(surv_obj ~ ScrArm, data = ERSPC)
```

```{r erspc-cox-cif, eval = eval_cs1, echo = FALSE, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model using \\code{survival::survfit} (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (1) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (2) Linear function of time which is equivalent to a Gompertz distribution. (3) The natural logarithm of time which is equivalent to a Weibull distribution. (4) Cubic B-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group")#,
  # ignore = 99
)

risks <- lapply(fits, function(i) {
  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
})

aft_fits <- lapply(list("exp","gompertz","weibull"), function(i)
  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm,
                           data = ERSPC, dist = i)
  )

names(aft_fits) <- list("exponential","gompertz","weibull")

# dev.off()
layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
     #                exp(coef(cox_model)), 
     #                exp(confint(cox_model))[1], 
     #                exp(confint(cox_model))[2]), 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Casebase)",
                  "Screening (Cox)", "Screening (Casebase)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = 15*0.97, y = 0.05/100, labels = "(3)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     xaxt = 'n', yaxt = 'n',
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(2)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(4)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

```{r, results='hide', eval = FALSE, echo = FALSE}
# par(mfrow = c(2, 2))
# lapply(fits, function(i) {
#   casebase::hazardPlot(i, newdata = new_data[2,], ci = FALSE)
#   casebase::hazardPlot(i, newdata = new_data[1,], ci = FALSE, add = TRUE, lty = 2)
# })
```

```{r erspc-casebase-exponential, eval = FALSE, echo = FALSE}
# # Equivalence between casebase and AFT
# # exponential -------------------------------------------------------------
# 
# casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
#                                                   data = ERSPC, 
#                                                   ratio = 100)
# aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
#                            data = ERSPC, dist = "exp")
# 
# -coef(aft_exponential)
# coef(casebase_exponential)
# summary(casebase_exponential)
# exp(coef(casebase_exponential)[2])
# exp(confint(casebase_exponential)[2, ])
# 
# # weibull -----------------------------------------------------------------
# 
# casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
#                                            data = ERSPC, 
#                                            ratio = 100)
# aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
#                        data = ERSPC, dist = "weibull")
# 
# -coef(aft_weibull)/aft_weibull$scale
# coef(casebase_weibull)
```

```{r erspc-casebase-cif, eval = FALSE, echo = FALSE}
# smooth_risk_exp <- casebase::absoluteRisk(
#   object = casebase_exponential,
#   time = seq(0, 15, 0.1),
#   newdata = new_data
# )
# 
# plot(survfit(cox_model, newdata = new_data),
#   xlab = "Years since Randomization",
#   ylab = "Cumulative Incidence (%)",
#   fun = "event",
#   xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
#   main = sprintf(
#     "Estimated Cumulative Incidence (risk) of Death from Prostate 
#                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
#     exp(coef(cox_model)),
#     exp(confint(cox_model))[1],
#     exp(confint(cox_model))[2]
#   ),
#   ylim = c(0, 2), yscale = 100
# )
# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 2], col = "red", lty = 2)
# lines(smooth_risk_exp[, 1], smooth_risk_exp[, 3], col = "blue", lty = 2)
# 
# 
# legend("topleft",
#   legend = c(
#     "Control group (Cox)", "Control group (Casebase)",
#     "Screening group (Cox)", "Screening group (Casebase)"
#   ),
#   col = c("red", "red", "blue", "blue"),
#   lty = c(1, 2, 1, 2),
#   bg = "gray90"
# )
```

<!-- The \code{fitSmoothHazard} function provides an estimate of the hazard function $h(x, t)$ is the hazard function, where $t$ denotes the numerical value of a point in prognostic/prospective time and $x$ is the realization of the vector $X$ of variates based on the patient's profile and intervention (if any). -->
<!-- The \code{absoluteRisk} function provides an estimate of the cumulative incidence curves for a specific risk profile using the following equation: -->
<!-- $$ CI(x, t) = 1 - \exp\left( - \int_0^t h(x, u) \textrm{d}u \right) $$ -->
